---
title: "Summer Temperature Regimes in NCBN Eelgrass Meadows"
execute:
  message: false
  warning: false
editor options: 
  markdown:
    wrap: none
output: 
  html_document:
    keep_md: false
    toc: true
    toc_float: true
    fig_caption: true
    number_sections: false
    df_print: kable
    theme: journal
    css: journalnps.min.css
    includes:
        before_body:
          - header.html
date: "`r format(Sys.time(), '%B %d, %Y')`"
params:
  park: "ASIS" #change to FIIS or CACO when knitting for park specific report
  park_name: "Cape Cod National Seashore"
---

```{=html}
<style>
p.caption {
  font-size: 0.8em;
}
</style>
```

```{r setup, include=FALSE}

# Libraries (load once)
library(tidyverse)
library(readxl)
library(ggrepel)
library(glue)
library(fs)
library(scales)   # for label_number
library(here) # use to load data from relative paths
library(knitr)
library(kableExtra)

options(knitr.table.format = "html")

# -------------------------------
# 0) GLOBAL STYLE + PALETTE
# -------------------------------
theme_set(theme_classic(base_size = 14))

# Locked legend order for Park_Code
park_levels <- c("ASIS", "FIIS", "CACO")

# Consistent colors and shapes everywhere
park_pal    <- c("ASIS"="#1f77b4","FIIS"="#ff7f00","CACO"="#2ca02c")
park_shapes <- c("ASIS"=16,"FIIS"=17,"CACO"=15)

# Accent colors for thresholds/extreme markers (not mapped to Park_Code)
grey_grid     <- "grey40"
accent_red    <- "#d62728"
accent_blue   <- "#1f78b4"
accent_purple <- "#8b3fb9"

# Reusable scales—use these in every plot that maps color/fill/shape to Park_Code
scale_park_color <- function(...) scale_color_manual(values = park_pal, breaks = park_levels, ...)
scale_park_fill  <- function(...) scale_fill_manual(values  = park_pal, breaks = park_levels, ...)
scale_park_shape <- function(...) scale_shape_manual(values = park_shapes, breaks = park_levels, ...)
```

```{r Load-Data, include=FALSE}
dat <- read_excel(here::here("data", "Site_Summary_Master_b.xlsx"), sheet = "Site_Year") %>%
  rename_with(~ gsub(" ", "_", .)) %>%
  mutate(
    Park_Code              = toupper(Park_Code),
    Park_Code              = factor(Park_Code, levels = park_levels), # lock legend order
    Mean_HDH_daily         = suppressWarnings(as.numeric(Mean_HDH_daily)),
    Mean_CDH_daily         = suppressWarnings(as.numeric(Mean_CDH_daily)),
    `%_abv_thres`          = suppressWarnings(as.numeric(`%_abv_thres`)),
    Number_warming_events  = suppressWarnings(as.numeric(Number_warming_events)),
    Year                   = suppressWarnings(as.integer(Year)),
    Transect               = toupper(gsub("[^A-Za-z]", "", trimws(Transect))),
    long_term_site = if_else(Site_Name %in% c("East Harbor", "Provincetown Harbor"), FALSE, TRUE)
  )


```

```{r helper-functions, include=FALSE}
# Site-year (avg across transects) -> Site-level (avg across years)
make_site_level <- function(df, metrics = c("Mean_HDH_daily","Mean_CDH_daily","%_abv_thres")) {
  df %>%
    dplyr::group_by(Park_Code, Site_Name, Year) %>%
    dplyr::summarise(
      HDH_year     = if ("Mean_HDH_daily" %in% metrics) mean(Mean_HDH_daily, na.rm = TRUE) else NA_real_,
      CDH_year     = if ("Mean_CDH_daily" %in% metrics) mean(Mean_CDH_daily, na.rm = TRUE) else NA_real_,
      pct_hot_year = if ("%_abv_thres"    %in% metrics) mean(`%_abv_thres`,    na.rm = TRUE) else NA_real_,
      .groups = "drop"
    ) %>%
    dplyr::group_by(Park_Code, Site_Name) %>%
    dplyr::summarise(
      mean_HDH = mean(HDH_year,     na.rm = TRUE),
      mean_CDH = mean(CDH_year,     na.rm = TRUE),
      pct_hot  = mean(pct_hot_year, na.rm = TRUE),
      n_years  = dplyr::n_distinct(Year),
      .groups  = "drop"
    ) %>%
    dplyr::mutate(Label = paste0(Site_Name, " (n=", n_years, ")"))
}

# Flag extreme years per site (e.g., 90th percentile of metric within each site)
flag_extreme_by_site <- function(df, value_col, prob = 0.9, flag_name = "Extreme") {
  df %>%
    dplyr::group_by(Site_Name) %>%
    dplyr::mutate(
      thresh = stats::quantile(.data[[value_col]], prob, na.rm = TRUE),
      "{flag_name}" := .data[[value_col]] >= .data[["thresh"]]
    ) %>%
    dplyr::ungroup()
  
}
```

```{r filter-to-park, include=FALSE}
park_code <- params$park
dat_pk <- dat %>% filter(Park_Code == park_code)
```

```{r Per-Park-Plotting, include=FALSE}
# A) High Heat Years (HDH) by site (facetted), one park
plot_hdh_extreme_by_site <- function(df, park_code) {
  site_year <- df %>%
    dplyr::filter(Park_Code == park_code) %>%
    dplyr::group_by(Site_Name, Year) %>%
    dplyr::summarise(Mean_HDH = mean(Mean_HDH_daily, na.rm = TRUE), .groups = "drop") %>%
    flag_extreme_by_site(value_col = "Mean_HDH", prob = 0.9, flag_name = "Extreme")
  
  ggplot(site_year, aes(x = Year, y = Mean_HDH)) +
    geom_line(color = "black", linewidth = 1) +
    geom_point(size = 2) +
    geom_point(data = subset(site_year, Extreme), color = accent_red, size = 3.5) +
    geom_text_repel(
      data = subset(site_year, Extreme),
      aes(label = Year), color = accent_red, size = 4, show.legend = FALSE,
      nudge_x = 0.4, nudge_y = 0.6, hjust = 0, direction = "y",
      box.padding = 0.3, point.padding = 0.2
    ) +
    geom_hline(aes(yintercept = thresh), linetype = "dashed", color = accent_red) +
    facet_wrap(~ Site_Name, ncol = 1, scales = "free_y") +
    scale_y_continuous(labels = label_number(accuracy = 1),
                       expand = expansion(mult = c(0.1, 0.2))) +
    scale_x_continuous(breaks = seq(min(site_year$Year, na.rm = TRUE), max(site_year$Year, na.rm = TRUE), by = 2)) +
    labs(
      title    = glue("Mean daily HDH by site ({park_code})"),
      subtitle = "Red dashed line = site-specific 90th percentile value; red points = extreme high heat years",
      x = "Year", y = "Mean daily heating degree hours"
    )
}

# B) Reduced Cooling Years (CDH) by site (facetted), one park
plot_cdh_lowcool_by_site <- function(df, park_code) {
  site_year <- df %>%
    dplyr::filter(Park_Code == park_code) %>%
    dplyr::group_by(Site_Name, Year) %>%
    dplyr::summarise(Mean_CDH = mean(Mean_CDH_daily, na.rm = TRUE), .groups = "drop") %>%
    flag_extreme_by_site(value_col = "Mean_CDH", prob = 0.9, flag_name = "Low_Cooling")
  
  ggplot(site_year, aes(x = Year, y = Mean_CDH)) +
    geom_line(color = "black", linewidth = 1) +
    geom_point(size = 2) +
    geom_point(data = subset(site_year, Low_Cooling), color = accent_blue, size = 3.5) +
    geom_text_repel(
      data = subset(site_year, Low_Cooling),
      aes(label = Year), color = accent_blue, size = 4, show.legend = FALSE,
      nudge_x = 0.4, nudge_y = 0.6, hjust = 0, direction = "y",
      box.padding = 0.3, point.padding = 0.2
    ) +
    geom_hline(aes(yintercept = thresh), linetype = "dashed", color = accent_blue) +
    facet_wrap(~ Site_Name, ncol = 1, scales = "free_y") +
    scale_y_continuous(labels = label_number(accuracy = 1),
                       expand = expansion(mult = c(0.1, 0.2))) +
    scale_x_continuous(breaks = seq(min(site_year$Year, na.rm = TRUE), max(site_year$Year, na.rm = TRUE), by = 2)) +
    labs(
      title    = glue("Mean daily CDH by site ({park_code})"),
      subtitle = "Blue dashed line = site-specific 90th percentile value; blue points = extreme low cooling years",
      x = "Year", y = "Mean daily cooling degree hours"
    )
}

# C) Double-Stress years (HDH high & CDH high), one park
plot_double_stress_by_site <- function(df, park_code) {
  site_year <- df %>%
    dplyr::filter(Park_Code == park_code) %>%
    dplyr::group_by(Site_Name, Year) %>%
    dplyr::summarise(
      Mean_HDH = mean(Mean_HDH_daily, na.rm = TRUE),
      Mean_CDH = mean(Mean_CDH_daily, na.rm = TRUE),
      .groups = "drop"
    ) %>%
    dplyr::group_by(Site_Name) %>%
    dplyr::mutate(
      hdh_thresh = stats::quantile(Mean_HDH, 0.9, na.rm = TRUE),
      cdh_thresh = stats::quantile(Mean_CDH, 0.9, na.rm = TRUE),
      High_Heat   = Mean_HDH >= hdh_thresh,
      Low_Cooling = Mean_CDH >= cdh_thresh,
      Double_Stress = High_Heat & Low_Cooling
    ) %>% dplyr::ungroup()
  
  ggplot(site_year, aes(x = Year, y = Mean_HDH)) +
    geom_line(color = "black", linewidth = 1) +
    geom_point(size = 2) +
    geom_point(data = subset(site_year, Double_Stress), color = accent_purple, size = 4) +
    geom_text_repel(
      data = subset(site_year, Double_Stress),
      aes(label = Year), color = accent_purple, size = 4, show.legend = FALSE,
      nudge_x = 0.4, nudge_y = 0.6, hjust = 0, direction = "y",
      box.padding = 0.3, point.padding = 0.2
    ) +
    { if(any(df$Double_Stress)) # turn off plotting of the HDH threshold line if there are no compound stress years
    geom_hline(aes(yintercept = hdh_thresh), linetype = "dashed", color = accent_red) } +
    facet_wrap(~ Site_Name, ncol = 1, scales = "free_y") +
    scale_y_continuous(labels = label_number(accuracy = 1),
                       expand = expansion(mult = c(0.1, 0.2))) +
    scale_x_continuous(breaks = seq(min(site_year$Year, na.rm = TRUE), max(site_year$Year, na.rm = TRUE), by = 2)) +
    labs(
      title    = glue("Compound stress years by site ({park_code})"),
      subtitle = "Purple points = years with extreme high heat AND low cooling",
      x = "Year", y = "Mean daily heating degree hours"
    )
}
```

```{r Regional-PLotting-Functions, include=FALSE}
# D) Site-level thermal stress across years (lines colored by Park_Code)
plot_sitelevel_thermal_stress_lines <- function(df) {
  site_year <- df %>%
    dplyr::filter(Park_Code %in% park_levels) %>%
    dplyr::group_by(Park_Code, Site_Name, Year) %>%
    dplyr::summarise(Mean_HDH = mean(Mean_HDH_daily, na.rm = TRUE), .groups = "drop")
  
  threshold <- stats::quantile(site_year$Mean_HDH, 0.9, na.rm = TRUE)
  
  site_labels <- site_year %>%
    dplyr::group_by(Site_Name) %>%
    dplyr::filter(Year == max(Year, na.rm = TRUE)) %>%
    dplyr::ungroup()
  
  ggplot(site_year,
         aes(x = Year, y = Mean_HDH, group = Site_Name, color = Park_Code)) +
    geom_line(alpha = 0.7, linewidth = 1) +
    geom_point(size = 2) +
    geom_text_repel(
      data = site_labels,
      aes(label = Site_Name),
      size = 3.5, direction = "y", hjust = 0,
      nudge_x = 1, box.padding = 0.5, point.padding = 0.3,
      segment.color = "grey60", show.legend = FALSE
    ) +
    geom_point(
      data = subset(site_year, Mean_HDH >= threshold),
      color = accent_red, size = 3
    ) +
    geom_text_repel(
      data = subset(site_year, Mean_HDH >= threshold),
      aes(label = Year),
      size = 3, color = accent_red, show.legend = FALSE
    ) +
    geom_hline(yintercept = threshold, linetype = "dashed", color = accent_red) +
    scale_park_color() +
    scale_x_continuous(breaks = seq(min(site_year$Year, na.rm = TRUE),
                                    max(site_year$Year, na.rm = TRUE), by = 2)) +
    expand_limits(x = max(site_year$Year, na.rm = TRUE) + 8) +
    labs(
      title    = "Site-Level Thermal Stress Across Years",
      subtitle = "Lines = sites; labels show site identity and extreme heat years",
      x = "Year", y = "Mean Heating Degree Hours per Day",
      color = "Park"
    )
}

# E) Thermal stress vs cooling (site-level scatter with size = % hot)
plot_stress_vs_cooling <- function(df) {
  site_level <- make_site_level(df, metrics = c("Mean_HDH_daily","Mean_CDH_daily","%_abv_thres"))
  # median_HDH <- 8.8
  # median_CDH <- -94.9
  # 
  # x_low  <- median_HDH * 0.5
  # x_high <- max(site_level$mean_HDH, na.rm = TRUE) * 0.8
  # y_low  <- min(site_level$mean_CDH, na.rm = TRUE) * 0.8
  # y_high <- median_CDH * 0.8
  
  cluster_df <- site_level %>% # create df with only Events and Exposure for fitting k-means
    mutate(id = paste0(Park_Code, "_", Site_Name)) %>%
    column_to_rownames(var = "id") %>%
    ungroup() %>%
    dplyr::select(mean_HDH, mean_CDH)
  
  km_fit <- kmeans(cluster_df, centers = 4, nstart = 2) # fit k-means clusters
  HDH_split <- mean(km_fit$centers[1:4])
  CDH_split <- mean(km_fit$centers[5:8])
  
  plot <- ggplot(site_level, aes(x = mean_HDH, y = mean_CDH, color = Park_Code, size = pct_hot)) +
    geom_point(alpha = 0.9) +
    geom_text_repel(aes(label = Label), size = 4) +
    geom_vline(xintercept = HDH_split, linetype = "dashed", color = grey_grid) +
    geom_hline(yintercept = CDH_split, linetype = "dashed", color = grey_grid) +
    # expand_limits(x = max(site_level$mean_HDH, na.rm = TRUE) * 1.2,
    #               y = min(site_level$mean_CDH, na.rm = TRUE) * 1.2) +
    scale_park_color() +
    guides(color = guide_legend(override.aes = list(alpha = 1, size = 4))) +
    labs(
      title = "Thermal Stress vs Cooling Relief Across Coastal Monitoring Sites",
      subtitle = "Site-level averages (transects averaged within year, then across years)",
      x = "Mean Heating Degree Hours per Day (Thermal Stress)",
      y = "Mean Cooling Degree Hours per Day (Thermal Recovery)",
      color = "Park",
      size  = "% Time Above Threshold"
    )
  
  plot_build <- ggplot_build(plot)
  x_limits <- plot_build$layout$panel_params[[1]]$x$continuous_range
  y_limits <- plot_build$layout$panel_params[[1]]$y$continuous_range
  
  plot +
    annotate("text", x = x_limits[1] + ((HDH_split - x_limits[1])/2), y = CDH_split + ((y_limits[2] - CDH_split)/2), vjust = 0.5, hjust = 0.5, label = "Stable Systems", size = 4.5, fontface = "bold") +
    annotate("text", x = x_limits[1] + ((HDH_split - x_limits[1])/2), y = y_limits[1] + ((CDH_split - y_limits[1])/2), vjust = 0.5, hjust = 0.5, label = "Thermal Refugia", size = 4.5, fontface = "bold") +
    annotate("text", x = HDH_split + ((x_limits[2] - HDH_split)/2), y = CDH_split + ((y_limits[2] - CDH_split)/2), vjust = 0.5, hjust = 0.5, label = "Chronic Heat Stress", size = 4.5, fontface = "bold") +
    annotate("text", x = HDH_split + ((x_limits[2] - HDH_split)/2), y = y_limits[1] + ((CDH_split - y_limits[1])/2), vjust = 0.5, hjust = 0.5, label = "Stress Buffered Systems", size = 4.5, fontface = "bold")
  
}

# F) Thermal balance (exposure vs cooling, site-level)
plot_thermal_balance <- function(df) {
  thermal_balance <- df %>%
    dplyr::group_by(Park_Code, Site_Name, Year) %>%
    dplyr::summarise(
      Exposure_year = mean(`%_abv_thres`, na.rm = TRUE),
      Cooling_year  = mean(Mean_CDH_daily,  na.rm = TRUE),
      .groups = "drop"
    ) %>%
    dplyr::group_by(Park_Code, Site_Name) %>%
    dplyr::summarise(
      MeanExposure = mean(Exposure_year, na.rm = TRUE),
      MeanCooling  = mean(Cooling_year,  na.rm = TRUE),
      n_years      = dplyr::n_distinct(Year),
      .groups      = "drop"
    ) %>%
    dplyr::mutate(Label = paste0(Site_Name, " (n=", n_years, ")"))
  
  ggplot(thermal_balance, aes(x = MeanExposure, y = MeanCooling, color = Park_Code)) +
    geom_point(size = 4, alpha = 0.9) +
    geom_text_repel(aes(label = Label), size = 3) +
    scale_park_color() +
    labs(
      title    = "Thermal Balance Across Coastal Monitoring Sites",
      subtitle = "Exposure vs cooling using consistent site-level averaging",
      x = "Mean % Time Above Threshold (Thermal Exposure)",
      y = "Mean Cooling Degree Hours (Thermal Recovery)",
      color = "Park"
    )
}

# G) Heat Event Structure (site-level; optional interpretive quadrants)
plot_heat_event_structure_site <- function(df) {
  site_level_events <- df %>%
    dplyr::group_by(Park_Code, Site_Name, Year) %>%
    dplyr::summarise(
      Events_year   = mean(Number_warming_events, na.rm = TRUE),
      Exposure_year = mean(`%_abv_thres`,        na.rm = TRUE),
      .groups = "drop"
    ) %>%
    dplyr::group_by(Park_Code, Site_Name) %>%
    dplyr::summarise(
      MeanEvents   = mean(Events_year,   na.rm = TRUE),
      MeanExposure = mean(Exposure_year, na.rm = TRUE),
      n_years      = dplyr::n_distinct(Year),
      .groups      = "drop"
    ) %>%
    dplyr::mutate(Label = paste0(Site_Name, " (n=", n_years, ")"))
  
  cluster_df <- site_level_events %>% # create df with only Events and Exposure for fitting k-means
    mutate(id = paste0(Park_Code, "_", Site_Name)) %>%
    column_to_rownames(var = "id") %>%
    ungroup() %>%
    dplyr::select(MeanEvents, MeanExposure)
  
  km_fit <- kmeans(cluster_df, centers = 4, nstart = 2) # fit k-means clusters
  event_split <- mean(km_fit$centers[1:4])
  exposure_split <- mean(km_fit$centers[5:8])
  
  plot <- ggplot(site_level_events, aes(x = MeanEvents, y = MeanExposure, color = Park_Code, shape = Park_Code)) +
    geom_point(size = 4.5) +
    geom_text_repel(aes(label = Label), size = 3, max.overlaps = 30) +
    geom_vline(xintercept = event_split,   linetype = "dashed", color = grey_grid) +
    geom_hline(yintercept = exposure_split,linetype = "dashed", color = grey_grid) +
    scale_park_color() +
    scale_park_shape() +
    labs(
      title    = "Heat Event Structure Across Coastal Parks",
      subtitle = "Points represent site-level averages across monitoring years",
      x = "Mean Number of Warming Events (Frequency)",
      y = "Mean % Time Above Threshold (Duration)",
      color = "Park", shape = "Park"
    )
  
  plot_build <- ggplot_build(plot)
  x_limits <- plot_build$layout$panel_params[[1]]$x$continuous_range
  y_limits <- plot_build$layout$panel_params[[1]]$y$continuous_range
  
  plot +
    annotate("text", x = x_limits[1] + ((event_split - x_limits[1])/2), y = exposure_split + ((y_limits[2] - exposure_split)/2), vjust = 0.5, hjust = 0.5, label = "Episodic Heat Waves", size = 4) +
    annotate("text", x = x_limits[1] + ((event_split - x_limits[1])/2), y = y_limits[1] + ((exposure_split - y_limits[1])/2), vjust = 0.5, hjust = 0.5, label = "Low Exposure", size = 4) +
    annotate("text", x = event_split + ((x_limits[2] - event_split)/2), y = exposure_split + ((y_limits[2] - exposure_split)/2), vjust = 0.5, hjust = 0.5, label = "Highly Stressed", size = 4) +
    annotate("text", x = event_split + ((x_limits[2] - event_split)/2), y = y_limits[1] + ((exposure_split - y_limits[1])/2), vjust = 0.5, hjust = 0.5, label = "Chronic Warming", size = 4)
}

# (Optional) H) Thermal balance at TRANSECT level (no interpretation grids)
plot_thermal_balance_transect <- function(df) {
  thermal_balance <- df %>%
    dplyr::filter(Park_Code %in% park_levels) %>%
    dplyr::group_by(Park_Code, Site_Name, Transect) %>%
    dplyr::summarise(
      MeanExposure = mean(`%_abv_thres`,    na.rm = TRUE),
      MeanCooling  = mean(Mean_CDH_daily,   na.rm = TRUE),
      n_years      = dplyr::n(),
      .groups      = "drop"
    ) %>%
    dplyr::mutate(Label = paste(Site_Name, Transect))
  
  ggplot(thermal_balance, aes(x = MeanExposure, y = MeanCooling, color = Park_Code)) +
    geom_point(size = 4) +
    geom_text_repel(aes(label = Label), size = 3) +
    scale_park_color() +
    labs(
      title = "Thermal Balance Across Parks (Transect Level)",
      x = "Mean % Time Above Threshold",
      y = "Mean Cooling Degree Hours",
      color = "Park"
    )
}

```

# Background

## Eelgrass Importance and Vulnerability

Eelgrass (*Zostera marina*) is a foundational marine plant that supports healthy coastal ecosystems. It improves water quality, stabilizes sediments, provides habitat for fish and invertebrates([Orth et. al 2006](#refs), [Hemminga and Duarte 2000](#refs), [Orth et al. 2020](#refs)), and functions as both a global and local carbon sink ([Sogin et al. 2022](#refs), [Fourqurean et al. 2012](#refs), [Novak et al. 2020](#refs)). Because eelgrass responds quickly to environmental change, it is often considered a “canary in the coal mine” for coastal health ([Orth et al. 2006](#refs)). Its ecological significance is one reason the Northeast Coastal and Barrier Network (NCBN) monitors seagrass condition over time, especially as concerns grow about the effects of warming waters. 

```{r, echo=FALSE, results='asis'}

if (params$park == "ASIS"| params$park=="FIIS") {
  cat("

Although *Zostera marina* is the dominant seagrass species in most Northeast Coastal and Barrier Parks, widgeon grass (*Ruppia maritima* ) also occurs at Assateague Island National Seashore (ASIS) and Fire Island National Seashore (FIIS). While *Ruppia* is not the primary focus of this analysis, it can respond differently to warming conditions, and studies have documented shifts from *Zostera* to *Ruppia* following warm summers, suggesting that rising temperatures may alter seagrass community composition over time ([Shields et al. 2019](#refs))

")
}
```


Globally, sea surface temperatures (SST) have increased significantly over the past century and are projected to continue rising in the 21st century ([IPCC, 2023](#refs)). In recent decades, eelgrass in park units along the Northeast U.S. coast has experienced water temperatures that approach or exceed known physiological limits ([Nagel et. al 2025a,b,c](#refs)). When water becomes too warm, eelgrass respiration outpaces photosynthesis, creating physiological stress that can lead to mortality; conversely, cooling periods allow these processes to rebalance, enabling plants to rebuild carbohydrate reserves needed for growth,biomass, and winter survival ([Lee et al. 2007](#refs), [Marín-Guirao et al. 2016](#refs), [Marsh et al. 1986](#refs)). Studies show that temperatures above 25 °C increase the likelihood of eelgrass mortality, and sustained exposure above 30 °C has caused complete die offs in areas such as Chesapeake Bay ([Jarvis & Moore 2010](#refs), [Novak et al. 2023](#refs)). Long term temperature records from New Hampshire to Maryland show that warmer than average summers are consistently followed by reduced eelgrass presence in the following year ([Plaisted et al. 2022](#refs)). When high temperature events persist without cooling relief, eelgrass loses the opportunity to recover from heat stress, causing stress to accumulate and increasing the risk of widespread decline ([Berger et al. 2024](#refs)).

These patterns of warmer summers, increased physiological stress, and fewer opportunities for recovery reflect broader climate driven trends. Marine heatwaves, periods when water temperatures exceed the 90^th^ percentile of a 30 year baseline, have become more frequent and intense ([Hobday etl al 2016](#refs)). Such events have already affected corals, kelp forests, and seagrasses worldwide ([Smale et al. 2019](#refs)). Research shows that eelgrass meadows experiencing both high heat and minimal cooling tend to lose shoot density and canopy height, while sites that receive periodic cooling recover more effectively between heat events ([Berger et al. 2024](#refs)).

Cumulative heat stress metrics are widely used to understand ecosystem and species-level impacts. NOAA’s Degree Heating Weeks (DHW) index, for example, tracks how often and for how long temperatures exceed thresholds that lead to coral bleaching and mortality ([Skirving et al. 2020](#refs)). These metrics emphasize that it is the accumulation of heat over time, not single temperature spikes, that often drive ecosystem decline. This same concept is directly relevant to eelgrass.

Characterizing the temperature regimes that influence eelgrass heat stress and recovery requires long-term temperature datasets. To better understand the thermal conditions shaping eelgrass meadow health within the Northeast Coastal and Barrier Network parks, we used long-term temperature datasets to:

1.  Identify years with extreme heat and limited cooling,
2.  Determine where these conditions overlapped to create compounded “stress years,” and
3.  Compare thermal regimes across eelgrass meadows in parks to understand regional vulnerability.

While temperature is a key driver of eelgrass stress and recovery, it operates alongside other environmental factors including water clarity, nutrients, salinity, dissolved oxygen, that together shape meadow condition ([Lefcheck et al. 2017](#refs), [Dunic and Cote, 2023](#refs), [Moore et al. 2012](#refs)). This report focuses on characterizing thermal regimes as a foundational step. Future analyses will explore how temperature interacts with these additional factors drive patterns in eelgrass condition across NCBN parks. 

# Methods

## Temperature Data Collection

The Northeast Coastal and Barrier Network uses the Tier-3 standardized Estuarine Nutrient Enrichment monitoring protocol ([Kopp and Neckles, 2009](#refs)) that includes seagrass condition monitoring (SOP 9). Seagrass condition is assessed using measures of population and shoot structure to track changes in meadow condition over time. The protocol was adapted from the international SeagrassNet program, which provides standardized methods for monitoring seagrass ecosystems.

Within the network, three coastal parks, Assateague Island National Seashore (ASIS), Cape Cod National Seashore, and Fire Island National Seashore (FIIS) implement the Tier-3 seagrass monitoring protocol. At `r params$park`, long-term seagrass condition is monitored at `r as.character(english::english(n_distinct(dat_pk$Site_Name[dat_pk$long_term_site == TRUE])))` site`r if (n_distinct(dat_pk$Site_Name) > 1) {'s'}`: `r knitr::combine_words(unique(dat_pk$Site_Name[dat_pk$long_term_site == TRUE]))`. `r if_else(params$park == "CACO", paste("Additionally, water temperature data was also collected at", knitr::combine_words(unique(dat_pk$Site_Name[dat_pk$long_term_site == FALSE])), "beginning in 2023 to provide broader ecological context, but these sites are not included in the long-term monitoring protocol."), "")`

At each long-term monitoring site, seagrass condition is monitored across three permanent 50-m transects that run parallel to shore at a shallow, mid-depth, and deep depth. `r if_else(params$park == "CACO", paste("Note that at East Harbor and Provincetown Harbor, temperature data was only collected at a single depth: XX meters.", ""), "")`Temperature loggers (HOBO UTBI-001 Tidbit Temp, Onset Computer Corporation) are attached to each transect and record water temperature at continuous 15-minute intervals throughout the year.

Following data collection, water temperature data from each year are uploaded to the NPS AQUARIUS WebPortal for QAQC, certification, and storage. Additional details about data collection at each site are shown in Table 1 below.

```{r Table1, echo=FALSE}

# Use your existing site_summary() and filter to the selected park
site_sum <- dat_pk %>%
  group_by(Park_Code, Site_Name) %>%
  summarise(
      n_years = dplyr::n_distinct(Year),
      first_year = min(Year, na.rm = TRUE),
      last_year = max(Year, na.rm = TRUE),
      is_longterm_site = if_else(any(long_term_site) == TRUE, "Yes", "No"),
      .groups = "drop"
    ) %>%
  arrange(Site_Name) %>%
  ungroup() %>%
  select("Site name" = Site_Name, "First year of data" = first_year, "Most recent year of data" = last_year, "Total years of data" = n_years, "Long-term monitoring site" = is_longterm_site) %>%
  {if (params$park != "CACO")
    select(., -c(`Long-term monitoring site`))
    else .}

# Clean, simple table using only knitr::kable()
kable(
  site_sum,
  caption = paste0(
    "Table 1. Site-specific details for temperature data available from each site in ",
    params$park,
    " after data filters were applied."
  ),
  booktabs = TRUE,
  label = NA,
  align = "c"
) %>%
  kable_styling(
    bootstrap_options = c("striped", "hover"),
    full_width = FALSE
    ) %>%
  row_spec(0, extra_css = "border-top: 2px solid lightgrey;") %>%
  row_spec(nrow(site_sum), extra_css = "border-bottom: 1px solid lightgrey")
```

## Data Summerization & Analysis

For this analysis, we focused on the continuous water temperature data and examined patterns of thermal stress and cooling dynamics with the goals of understanding the thermal conditions at `r params$park` and comparing with those at other long-term monitoring sites at `r knitr::combine_words(unique(dat %>% filter(Park_Code != params$park) %>% pull(Park_Code)))`.

Prior to analyses, temperature data were first filtered to the months from May-September to hone in on when thermal stress likely occurs. To ensure that the data from each sampling year occurred consistently over the same time periods, we then further filtered the data from each site to only include years with 80% data coverage and with at least 10 days of data in a minimum of three of the five focal months. Then, data from the individual transects at each site were summarized up to a single thermal metric value (HDH or CDH) for each year as described in the next section.

### Novel Thermal Metrics for Eelgrass

For eelgrass, prolonged exposure to high temperatures can lead to physiological stress where respiration outpaces photosynthesis. To describe long-term thermal conditions in eelgrass meadows, we calculated two site-level yearly metrics from the continuous water temperature data: **heating degree hours (HDH)** and **cooling degree hours (CDH)** ([Berger et al. 2024](#refs)). HDH is calculated by first summing the total number of hours above a park-specific temperature threshold across the focal months (May to September) to get the total HDH per year. The total HDH values are then divided by the sum of all hours in the period of record for each year to get the mean HDH per year. Similarly, mean CDH for each year is calculated using the same method but for the number of hours below the park-specific temperature threshold.

Park-specific temperature thresholds were calculated from the filtered water temperature data described in the previous section. We first calculated the mean monthly temperature for each month between May to September across all years for each transect at each site (i.e., transect-level mean monthly temperatures). Then, we calculated the maximum of the transect-level mean monthly temperatures for each site. Note that we choose to use maximum temperatures because the distribution of seagrasses (and many other vegetation communities - such as mangroves) have been shown to be controlled in part by the frequency and intensity of climate extremes (citations), which are also likely to be altered by climate change (citations). Finally, we averaged the maximum mean monthly temperature from the site with the lowest value and the site with the highest value within each park to get park-specific temperature threshold values. We calculated park-specific temperature thresholds of 24.6 °C for CACO, 26.3°C for FIIS, and 28.4°C for ASIS. Notably, these park-specific temperature thresholds are extremely similar to known values of eelgrass thermal tolerance reported in peer-reviewed literature (e.g., [Berger & Berg 2024](https://doi.org/10.3354/meps14587), [Moore & Jarvis 2008](https://doi.org/10.2112/SI55-014), [Carr et al. 2012](https://doi.org/10.3354/meps09556), [Shields et al. 2018](https://doi.org/10.3390/d10040125)).

Thus, these metrics capture not just how hot or cold the water gets, but how long water temperatures remain above or below a biological threshold. When examined together, HDH and CDH help to understand the factors controlling eelgrass loss and recovery across coastal National Parks in the northeast (Figure 1).

```{r concepts, echo=FALSE, warning=FALSE, fig.show='hold', fig.width=4, fig.height=3, fig.cap="Figure 1. Conceptualized diagrams demonstrating how yearly heating degree hours (HDH) and cooling degree hours (CDH) are calculated for each site."}
### HDH and CDH Concept##

threshold <- if_else(params$park == "ASIS", 28.5, 25)

df <- tibble(
  time = 1:100,
  temp = threshold + sin(time/10)*3
)

ggplot(df, aes(x = time, y = temp)) +
  geom_line(size = 1.2) +
  geom_hline(yintercept = threshold, linetype = "dashed") +
  geom_ribbon(
    data = df %>% filter(temp > threshold),
    aes(ymin = threshold, ymax = temp),
    fill = "red",
    alpha = 0.4
  ) +
  labs(title = "HDH = Heat Above Threshold") +
  xlab("Percent Time (%)") +
  ylab("Temperature (°C)") +
  theme_classic()

ggplot(df, aes(x = time, y = temp)) +
  geom_line(size = 1.2) +
  geom_hline(yintercept = threshold, linetype = "dashed") +
  geom_ribbon(
    data = df %>% filter(temp < threshold),
    aes(ymin = temp, ymax = threshold),
    fill = "blue",
    alpha = 0.4
  ) +
  labs(title = "CDH = Cooling Below Threshold") +
  xlab("Percent Time (%)") +
  ylab("Temperature (°C)") +
  theme_classic()


```

### Site-Specific Extremes

Within each site at `r params$park`, we identified extreme high heat years as those years where the mean HDH was above the site-specific 90^th^ percentile (`r paste0(threshold, "°C")`). These years represent years when the eelgrass meadow at these sites experienced the most intense and prolonged thermal stress. Similarly, we also identified extreme low cooling years as those years where the mean CDH was above the 90^th^ percentile for each site. In this context, lower CDH values (i.e., more negative) represent greater cooling below the temperature threshold at each site, indicating potential periods of relief for eelgrass meadows. Finally, we identified compound stress years as those years with both extreme high heat and extreme low cooling.

### Regional Comparisons

In order to understand regional patterns of seagrass thermal stress and cooling relief among coastal parks in the northeast, we used k-means cluster analysis to categorize each site in ASIS, CACO, and FIIS into four thermal stress regimes and four cooling relief regimes. K-means cluster analyses were conducted using 4 centers (to represent the 4 thermal stress or cooling relief regimes) and 2 random sets via the algorithm defined by [Hartigan and Wong](#refs) as implemented in the 'kmeans' function from the 'stats' package in R.

The four thermal stress regimes were based on 1) the % time above the previously mentioned park-specific threshold averaged across all years (i.e., duration of warming), and 2) the number of warming events averaged across all years (i.e., frequency of warming). Thermal stress regimes were defined as:

-   **Episodic Heatwave**: Sites with infrequent, but intense heat events
-   **Highly Stressed**: Sites with frequent and prolonged heat events
-   **Low Exposure**: Sites with infrequent and limited time above thermal threshold
-   **Chronic Warming**: Sites with frequent heat events that are not prolonged (pulsing)

The four cooling relief regimes were based on 1) the mean cooling degree hours averaged across all years at each site, and 2) the mean heating degree hours averaged across all years at each site. Cooling relief regimes were defined as:

-   **Stable Systems**: Low heat stress and moderate cooling
-   **Chronic Heat Stress**: High heat stress and limited cooling
-   **Thermal Refugia**: Low heat stress and strong cooling
-   **Stress Buffered Systems**: High heat stress but strong cooling

# Results

## Extreme High Heat Years

```{r extreme-HDH-years, echo=FALSE}
# Prepare site-year data for the selected park
site_year <- dat %>%
  filter(Park_Code == params$park) %>%
  group_by(Site_Name, Year) %>%
  summarise(Mean_HDH = mean(Mean_HDH_daily, na.rm = TRUE), .groups = "drop")

# Use your helper to flag extreme years (90th percentile)
site_year_extreme <- flag_extreme_by_site(site_year, value_col = "Mean_HDH", prob = 0.9, flag_name = "Extreme") %>%
  filter(Extreme) %>%
  arrange(Site_Name, Year)

site_result_text <- function(site_year_extreme_df, metric, park_code) {
  
  sites <- unique(dat_pk$Site_Name)
  
  extreme_sites <- site_year_extreme_df %>%
    {if (metric == "HDH" | metric == "CDH")
      filter(., Extreme)
      else if (metric == "double_stress")
        filter(., Double_Stress)
    } %>%
    group_by(Site_Name) %>%
    summarise(years = as.character(knitr::combine_words(Year))) %>%
    mutate(text = paste(years, "for", Site_Name))
  
  no_extreme_sites <- sites[(!sites %in% extreme_sites$Site_Name)]
  
  extreme_sites_text <- if (metric == "HDH") {
    paste0("Extreme high heat years were ", knitr::combine_words(extreme_sites$text), " (Figure 2).")
  } else if (metric == "CDH") {
    paste0("Extreme low cooling years were ", knitr::combine_words(extreme_sites$text), " (Figure 3).")
  } else if (metric == "double_stress") {
    paste0("Compound stress years were ", knitr::combine_words(extreme_sites$text), " (Figure 4).")
  }
  
  all_sites_text <- case_when(
    length(no_extreme_sites) == 0 ~ extreme_sites_text,
    
    length(no_extreme_sites) >= 1 & length(no_extreme_sites) < length(sites) & metric == "HDH" ~ paste0(extreme_sites_text, " There were no extreme high heat years at ", knitr::combine_words(no_extreme_sites, and = " or "), "."),
    length(no_extreme_sites) >= 1 & length(no_extreme_sites) < length(sites) & metric == "CDH" ~ paste0(extreme_sites_text, " There were no extreme low cooling years at ", knitr::combine_words(no_extreme_sites, and = " or "), "."),
    length(no_extreme_sites) >= 1 & length(no_extreme_sites) < length(sites) & metric == "double_stress" ~ paste0(extreme_sites_text, " There were no compound stress years at ", knitr::combine_words(no_extreme_sites, and = " or "), "."),
    
    length(no_extreme_sites) == length(sites) & metric == "HDH" ~ paste0("There were no extreme high heat years at any of the sites at ", park_code, " (Figure 2)."),
    length(no_extreme_sites) == length(sites) & metric == "CDH" ~ paste0("There were no extreme low cooling years at any of the sites at ", park_code, " (Figure 3)."),
    length(no_extreme_sites) == length(sites) & metric == "double_stress" ~ paste0("There were no compound stress years at any of the sites at ", park_code, " (Figure 4).")
  )
  
  return(all_sites_text)
}
```

`r site_result_text(site_year_extreme, "HDH", params$park)`

```{r plot-hdh, fig.width=10, fig.height=8, fig.cap=paste0("Figure 2. Mean yearly HDH for each site at ", params$park, ". Extreme high heat years (red points) are years when mean HDH exceeded the site-specific 90th percentile value (red dashed line)."), echo=FALSE}
plot_hdh_extreme_by_site(dat, park_code)
```

## Extreme Low Cooling Years

```{r extreme-CDH-years, echo=FALSE}

# Prepare site-year data for the selected park
site_year_cdh <- dat %>%
  filter(Park_Code == params$park) %>%
  group_by(Site_Name, Year) %>%
  summarise(Mean_CDH = mean(Mean_CDH_daily, na.rm = TRUE), .groups = "drop")

# Use your helper to flag extreme low cooling years (90th percentile)
site_year_cdh_extreme <- flag_extreme_by_site(site_year_cdh, value_col = "Mean_CDH", prob = 0.9, flag_name = "Extreme") %>%
  filter(Extreme) %>%
  arrange(Site_Name, Year)
```

`r site_result_text(site_year_cdh_extreme, "CDH", params$park)`

```{r plot-cdh, fig.width=10, fig.height=8, fig.cap=paste0("Figure 3. Mean yearly CDH for each site at ", params$park, ". Extreme low cooling years (blue points) are years when mean CDH exceeded the site-specific 90th percentile value (blue dashed line)."), echo=FALSE}
plot_cdh_lowcool_by_site(dat, park_code)
```

## Compound Stress Years

```{r double-stress-years, echo=FALSE}

# Summarize double‑stress years (HDH high & CDH high) for each site in the selected park

double_stress <- dat %>%
  filter(Park_Code == params$park) %>%
  group_by(Site_Name, Year) %>%
  summarise(
    Mean_HDH = mean(Mean_HDH_daily, na.rm = TRUE),
    Mean_CDH = mean(Mean_CDH_daily, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  group_by(Site_Name) %>%
  mutate(
    hdh_thresh = quantile(Mean_HDH, 0.9, na.rm = TRUE),
    cdh_thresh = quantile(Mean_CDH, 0.9, na.rm = TRUE),
    High_Heat   = Mean_HDH >= hdh_thresh,
    Low_Cooling = Mean_CDH >= cdh_thresh,   # high CDH = low cooling
    Double_Stress = High_Heat & Low_Cooling
  ) %>%
  arrange(Site_Name, Year)
```

`r site_result_text(double_stress, "double_stress", params$park)`

```{r plot-stress, fig.width=10, fig.height=8, fig.cap="Figure 4. Compound stress years (purple points) are years when mean CDH **and** mean HDH exceeded the site-specific 90th percentile values.", echo=FALSE}
plot_double_stress_by_site(dat,park_code)
```

## How does `r params$park` compare across the region?

```{r regional-results, echo=FALSE, fig.width=10, fig.height=6, fig.cap="Figure 5. Heat event structure across three coastal National Park Service units (CACO, ASIS, FIIS). Each point represents a site-level average of the mean number of warming events per year (frequency) and the mean percentage of time above the thermal threshold (duration). Dashed lines divide the figure into four interpretive thermal regimes. Labels show the average number of monitoring years contributing to each site-level estimate."}
plot_heat_event_structure_site(dat)
```

Across the three coastal NCBN parks, clear regional patterns emerge. At CACO, most sites fall in the category of having fewer and shorter heat‑stress events. An exception is the partially restored East Harbor system, which shows a different pattern. Overall, CACO sites tend to experience heat stress as brief or infrequent pulses rather than long, sustained events.

In contrast, Tingles Island at ASIS experiences heat stress much more often, and these events add up throughout the season. This pattern is expected for parks located in warmer parts of the region, where longer and more frequent heat events are common.

FIIS sites fall between these two extremes. Heat‑stress conditions vary widely across the park, depending on where each site is located. For example, Moriches Bay shows short, frequent warming events, while Great South Bay experiences longer or more sustained heat episodes. This mix reflects the diversity of habitats and local conditions across Fire Island.

These results show that sites across the regional scale experience different heat stress conditions and even sites within the same park respond in different ways. This raises an important questions that will require further data collection and analysis: **how are some of the most chronically stressed sites, like Moriches Bay, Pleasant Bay, and Tingles Island still able to support eelgrass populations?**

## What are the impacts of cooling?

Now that we've seen sites experience heat stress differently, we can look at cooling relief from heat events. Here, we divide the eelgrass into the four previously described cooling regimes based on the balance between thermal stress and thermal relief:

```{r, echo=FALSE, fig.width=12, fig.height=6, fig.cap="Figure 6. Thermal stress and cooling recovery across three coastal National Park Service units (CACO, ASIS, FIIS). Each point represents a site‑level average of the mean heating degree hours per day (thermal stress) and the mean cooling degree hours per day (thermal recovery). Dashed lines divide the figure into four interpretive recovery regimes. Labels show the average number of monitoring years contributing to each site‑level estimate."}
plot_stress_vs_cooling(dat)
```

The balance between heat stress events and cooling recovery also varies across the region and within parks. Pleasant Bay and Moriches Bay receive some of the strongest cooling periods among the chronically warm sites, second only to the fully open tidal systems on Cape Cod. This pattern reflects their position relative to inlets where tidal flushing brings in cooler ocean water that helps moderate warmer temperatures and sustain populations of eelgrass. In contrast, sites with more limited flushing, like Great South Bay and East Harbor, show weaker cooling relief consistent with restricted or partially restored flow. These sites match site conditions that no longer support eelgrass. Tingles Island falls between these two outcomes as it receives little cooling, sits between a stable and thermal refugia zone, and supports sparse eelgrass populations.

# Key Takeaways

```{r, echo=FALSE, results='asis'}

if (params$park == "ASIS") {
  cat("
### ASIS Key Takeaways

Across NCBN parks, Tingles Island stands out as one of the most heat-stressed locations due to frequent and prolonged heat events that accumulate throughout the season. This thermal pattern is expected given its southern latitude relative to FIIS (NY) and CACO (MA).  The Tingles Island transect supports only sparse eelgrass populations, reflecting moderate cooling, but still challenging environment compared to denser sites like Pleasant Bay and Moriches Bay. 

While this analysis focuses on thermal conditions alone, other water quality factors like dissolved oxygen, turbidity, salinity, and pH also play a key role in shaping eelgrass health. At Tingles Island, long- term trend reports show significant declines in eelgrass conditions (canopy height, biomass, and percent cover, shoot density, and biomass ratios), trends driven by a combination of chronically poor water quality and heat stress (Nagel et al. 2025a). Together these observations point to ecological pressure on the system.

")
}


# --- FIIS paragraph ---
if (params$park == "FIIS") {
  cat("
### FIIS Key Takeaways

Across NCBN parks, Great South Bay is one of the most heat-stressed locations experiencing frequent and prolonged heat events that accumulate throughout the season. This SeagrassNet site currently supports no eelgrass, while nearby Moriches Bay maintains dense eelgrass beds and experiences shorter heat waves due to its greater cooling capacity. 

While this analysis focuses on thermal condition alone, other water quality factors like dissolved oxygen, turbidity, salinity, and pH also play a key role in shaping eelgrass health. At Great South Bay, long term trend reports show a stark decline in eelgrass condition across all three transects from 2007 to 2011 (Nagel et. al 2025c). This decline is likely driven by poor water clarity combined with prolonged summer temperatures exceeding 25 degrees Celsius. Together these conditions point to ecological pressure on the system. 

")
}

## ---CACO paragraph ---
if (params$park == "CACO") {
  cat("
### CACO Key Takeaways

Across Cape Cod, eelgrass sites span a wide range of thermal regimes driven by local hydrodynamics. These range from open coast settings with relatively low heat stress (Duck Harbor and Provincetown Harbor), to coastal lagoons like Pleasant Bay, and highly restricted systems such as East Harbor. **East Harbor is the most heat stressed NCBN site**, experiencing infrequent but intense heat events that far exceed typical seasonal conditions. This site currently supports no eelgrass, while Pleasant Bay and Provincetown Harbor maintain eelgrass beds and benefit from stronger cooling that limits the duration of heat waves. 

While this analysis focuses on thermal condition alone, other water quality factors like dissolved oxygen, turbidity, salinity, and pH also play a key role in shaping eelgrass health. At Duck Harbor, long term reports shows a marked decline in eelgrass condition across the shallow and deep transects from 2014 to 2022 (Nagel et al. 2025c). In Pleasant Bay, eelgrass persists only on the mid depth transect, with losses from the shallow and deep transects and an overall declining trend. These patterns are likely driven by a combination of wave energy (erosion or burial), reduced water clarity, and prolonged summer temperatures exceeding 25C.These patterns highlight the varying vulnerabilities and site dynamics across CACO sites.
")
}
```

In parallel with these environmental assessments, NPS scientists and partners are using genomics to better understand why some eelgrass populations can withstand warming while other decline. By sequencing plants from Maine to North Carolina, we can identify genetic differences linked to heat tolerance and local adaptation. These results will guide future restoration planning and identify donor populations for transplanting seeds to help vulnerable meadows keep pace with rising sea temperatures.

> **🟦 Network‑Wide Key Takeaways**
>
> • Across NCBN coastal parks, eelgrass beds experience a wide range of thermal conditions, from sites with prolonged, cumulative heat exposure to sites with strong cooling capacity.
>
> • Understanding both heating intensity and cooling relief provides essential context for interpreting long‑term eelgrass trends, genomic thermal tolerance, and vulnerability to climate‑driven change.
>
> • Applying thermal metrics alongside genomic data will help identify populations with potential heat tolerance and strengthen site‑selection models for restoration and transplanting.
>
> • Integrating eelgrass distribution from long‑term transects with temperature patterns will help parks diagnose change, identify drivers of decline, and prioritize management actions.

*These thermal metrics are part of an ongoing collaborative project and represent a novel approach to characterizing eelgrass stress across NCBN parks. Thresholds linking these metrics to genomic patterns and adaptive capacity are still being refined in coordination with project cooperators. Readers can learn more about the broader genomics effort, restoration framework, and supporting documentation on the [Resilient Eelgrass Project Sharepoint](https://doimspp.sharepoint.com/:u:/r/sites/nps-ncbn/SitePages/Seagrass-Resilience-%26-Restoration.aspx?csf=1&web=1&e=nbw5PA).*

# References

Berger, A.C. & Berg, P. (2024) Eelgrass meadow response to heat stress. I. Temperature threshold for ecosystem production derived from in situ aquatic eddy covariance measurements. Marine Ecology Progress Series 736:35-46. <https://doi.org/10.3354/meps14587>

Berger A.C., Berg P., McGlathery K.J., Aoki L.R., & Kerns K. (2024) Eelgrass meadow response to heat stress. II. Impacts of ocean warming and marine heatwaves measured by novel metrics. Marine Ecology Progress Series 736:47-62. <https://doi.org/10.3354/meps14588>

Dunic, J. C., & Côté, I. M. (2023). Management thresholds shift under the influence of multiple stressors: eelgrass meadows as a case study. Conservation Letters, 16(2), e12938.

Fourqurean, J. W., Duarte, C. M., Kennedy, H., Marbà, N., Holmer, M., Mateo, M. A., ... & Serrano, O. (2012). Seagrass ecosystems as a globally significant carbon stock. Nature geoscience, 5(7), 505-509.

Hartigan, J.A. & Wong, M.A. (1979) Algorithm AS 136: A k-means clustering algorithm. Applied Statistics 28:100-108. <https://doi.org/10.2307/2346830>

Hobday AJ, Alexander LV, Perkins SE, Smale DA and others. (2016) A hierarchical approach to defining marine heatwaves. Prog Oceanogr 141: 227– 238. <https://doi.org/10.1016/j.pocean.2015.12.014>

Hemminga, M. A., & Duarte, C. M. (2000). Seagrass ecology. Cambridge University Press.

IPCC, 2023: Climate Change 2023: Synthesis Report. Contribution of Working Groups I, II and III to the Sixth Assessment Report of the Intergovernmental Panel on Climate Change [Core Writing Team, H. Lee and J. Romero (eds.)]. IPCC, Geneva, Switzerland, 184 pp. https://doi.org/10.59327/IPCC/AR6-9789291691647.

Jarvis, J.C. & Moore, K.A. (2010) The role of the photosynthetic pathway in the acclimation of Zostera marina to light and temperature stress. Journal of Experimental Marine Biology and Ecology 395:105–110.

Kopp, B.S., & Neckles, H.A. (2009) A protocol for monitoring estuarine nutrient enrichment in coastal parks of the National Park Service Northeast Region. Natural Resource Report NPS/NCBN/NRR—2009/110. National Park Service, Fort Collins, Colorado. <https://irma.nps.gov/DataStore/Reference/Profile/661689>

Lefcheck, J. S., Wilcox, D. J., Murphy, R. R., Marion, S. R., & Orth, R. J. (2017). Multiple stressors threaten the imperiled coastal foundation species eelgrass (Zostera marina) in Chesapeake Bay, USA. Global Change Biology, 23(9), 3474-3483.

Marín-Guirao, L., Ruiz, J. M., Dattolo, E., Garcia-Munoz, R., & Procaccini, G. (2016). Physiological and molecular evidence of differential short-term heat tolerance in Mediterranean seagrasses. Scientific reports, 6(1), 28615.

Marsh Jr, J. A., Dennison, W. C., & Alberte, R. S. (1986). Effects of temperature on photosynthesis and respiration in eelgrass (Zostera marina L.). Journal of Experimental Marine Biology and Ecology, 101(3), 257-267.

Nagel, J. L., H. K. Plaisted, C. P. Peck, and P. Duffy. 2025a. Condition and trends of estuarine water quality and seagrass in Assateague Island National Seashore: Northeast Coastal and Barrier Network, 2005–2022. Science Report NPS/SR—2025/273. National Park Service, Fort Collins, Colorado. <https://doi.org/10.36967/2310561>

Nagel, J. L., H. K. Plaisted, C. P. Peck, and P. Duffy. 2025b. Condition and trends of estuarine water quality and seagrass in Cape Cod National Seashore: Northeast Coastal and Barrier Network, 2003–2022. Science Report NPS/SR—2025/267. National Park Service, Fort Collins, Colorado. \< <https://doi.org/10.36967/2310108>

Nagel, J. L., H. K. Plaisted, C. P. Peck, and P. Duffy. 2025c. Condition and trends of estuarine water quality and seagrass in Fire Island National Seashore: Northeast Coastal and Barrier Network, 2003–2022. Science Report NPS/SR—2025/294. National Park Service, Fort Collins, Colorado. <https://doi.org/10.36967/2311455>

Novak, A. B., Pelletier, M. C., Colarusso, P., Simpson, J., Gutierrez, M. N., Arias-Ortiz, A., ... & Vella, P. (2020). Factors influencing carbon stocks and accumulation rates in eelgrass meadows across New England, USA. Estuaries and Coasts, 43(8), 2076-2091.

Novak, A.B., Plaisted, H.K., Hughes, Z.J., Mittermayr, A., & Molden, E. (2023) Eelgrass (<i>Zostera marina L.</i>) populations are threatened by high sea-surface temperatures and impaired waters on Nantucket Island, USA. Marine Pollution Bulletin 197:115689. <https://doi.org/10.1016/j.marpolbul.2023.115689>

Moore, K. A., Shields, E. C., Parrish, D. B., & Orth, R. J. (2012). Eelgrass survival in two contrasting systems: role of turbidity and summer water temperatures. Marine Ecology Progress Series, 448, 247-258.

Orth, R.J., Carruthers, T.J., Dennison, W.C., Duarte, C.M., Fourqurean, J.W., Heck, K.L., Hughes, A.R., Kendrick, G.A., Kenworthy, W.J., Olyarnik, S., Short, F.T., Waycott, M. & Williams, S. L. (2006). A global crisis for seagrass ecosystems. BioScience 56:987-996. [https://doi.org/10.1641/0006-3568(2006)56[987:AGCFSE]2.0.CO;2](https://doi.org/10.1641/0006-3568(2006)56%5B987:AGCFSE%5D2.0.CO;2){.uri}

Orth, R. J., Lefcheck, J. S., McGlathery, K. S., Aoki, L., Luckenbach, M. W., Moore, K. A., ... & Lusk, B. (2020). Restoration of seagrass habitat leads to rapid recovery of coastal ecosystem services. Science advances, 6(41), eabc6434.

Plaisted, H.K., Shields, E.C., Novak, A.B., Peck, C.P., Schenck, F., Carr, J., Duffy, P.A., Evans, N.T., Fox, S.E., Heck, S.M., Hudson, R., Mattera, T., Moore, K.A., Neikirk, B., Parrish, D.B., Peterson, B.J., Short, F.T. & Tinoco, A.I. (2022) Influence of rising water temperature on the temperate seagrass species eelgrass (<i>Zostera marina L.</i>) in the Northeast USA. Frontiers in Marine Science 9:920699. <https://doi.org/10.3389/fmars.2022.920699>

Shields, E. C., Parrish, D., & Moore, K. (2019). Short-term temperature stress results in seagrass community shift in a temperate estuary. Estuaries and Coasts, 42(3), 755-764.

Skirving, W., Marsh, B., De La Cour, J., Liu, G., Harris, A., Maturi, E., Geiger, E. & Eakin, C.M. (2020) CoralTemp and the coral reef watch coral bleaching heat stress product suite version 3.1. Remote Sensing 12:3856. <https://doi.org/10.3390/rs12233856>

Smale, D. A., Wernberg, T., Oliver, E. C., Thomsen, M., Harvey, B. P., Straub, S. C., ... & Moore, P. J. (2019). Marine heatwaves threaten global biodiversity and the provision of ecosystem services. Nature Climate Change, 9(4), 306-312.

Sogin, E. M., Michellod, D., Gruber-Vodicka, H. R., Bourceau, P., Geier, B., Meier, D. V., ... & Liebeke, M. (2022). Sugars dominate the seagrass rhizosphere. Nature Ecology & Evolution, 6(7), 866-877.

## Contact Information

| Katie Button, NCBN Biologist
| Kaitlyn_Button\@nps.gov
| 
| Holly Plaisted, NCBN Biologist & Seagrass Protocol Lead
| Holly_Plaisted\@nps.gov
| 
| Laura Feher, NCBN Biologist & Data Manager
| laura_feher\@nps.gov
| 
| [Vital Signs: Northeastern Coastal & Barrier Inventory and Monitoring Network](https://www.nps.gov/im/ncbn/what-we-monitor.htm)
| [Reports & Publications (U.S. National Park Service) (nps.gov)](https://www.nps.gov/im/ncbn/reports-publications-data.htm)
